Astrometry of the Neptune-Triton System

### 2017-08-16 relatorio/Netuno/Relatorio_2.tex
mod 2017 ago16

\documentclass[12pt,a4paper]{report}
\usepackage[brazilian, english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amsthm,amsfonts,amssymb,textcomp}
%\usepackage{latexsym}
\usepackage{graphicx}
\graphicspath{{figuras}}
\usepackage{subfigure}
\usepackage{float}
\usepackage{longtable}
\usepackage{color}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage[breaklinks=true]{hyperref}
\usepackage[comma,authoryear]{natbib}
\usepackage[nonumberlist]{glossaries}
\usepackage{arydshln}
\usepackage{footnote}
\usepackage{longtable}
\usepackage[small,bf,singlelinecheck=off]{caption}
\usepackage[left=3cm,right=2cm,top=3cm,bottom=2cm]{geometry}
\usepackage{arydshln}
\usepackage{scalefnt}
%\usepackage[alf]{abntcite}

%%% newcommand %%%%%%%%%%%%%%%%%
\newcommand{\PE}{Perkin-Elmer }
\newcommand{\BC}{Boller \& Chivens }
\newcommand{\noccs}{5442 }

@@

%\input{glossario}
\author{Altair Ramos Gomes Júnior}
\title{Occultation Predictions of Irregular Satellites}
\begin{document}

@@

\section*{Introduction}\label{Sec: introducao}

Irregular satellites revolve around giant planets at large distances in eccentric, highly inclined and frequently retrograde orbits. Because of these peculiar orbits, it is largely accepted that these objects did not form by accretion around their planet, but were captured in the early Solar System \citep{Sheppard2005}.

There is no consensus for a single model explaining where the irregular satellites were formed. \cite{Cuk2004} showed that the progenitor of the Himalia group may have originated in heliocentric orbits similar to the Hilda asteroid group. \cite{Sheppard2005} stated that the irregular satellites may be some of the objects that were formed within the giant planets region.

\cite{Grav2003} and \cite{Grav2007} showed that the irregular satellites from the giant planets have their colors and spectral slopes similar to C-, D- and P-type asteroids, Centaurs and trans-neptunian objects (TNOs). This suggests that they may have come from different locations in the early solar system.

\cite{Sheppard2005} and \cite{Jewitt2007} also explored the possibility that the irregular satellites originated as comets or TNOs. TNOs are highly interesting objects that, due to their large heliocentric distances, may be highly preserved with physical properties similar to those they had when they were formed \citep{Barucci2008}. This is even more true for the smaller objects, since in principle larger sizes favour physical differentiation processes in the body and vice-versa. However, due to the distance, the smaller TNOs from this region are more difficult to observe. Thus, if irregular satellites - or at least a few of them - do share a common origin with small TNOs, and since these objects are situated at much closer heliocentric distances now, this gives a unique chance of observing and studying representatives of this specific TNO population in much greater detail than could ever be possible by direct observation of this population in the Kuiper Belt.

Phoebe is the most studied irregular satellite. \cite{Clark2005} suggest that its surface is probably covered by material of cometary origin. It was also stated by \cite{Johnson2005} that if the porosity of Phoebe is 15\%, Phoebe would have an uncompressed density similar to those of Pluto and Triton.

Size, shape, albedo and composition would help a lot to trace back their true origin, but these physical parameters are yet poorly known for irregular satellites. The observation of stellar occultations would allow for the determination of such parameters. Indeed Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018, improving a lot the chances of observing such events in the near future.

%\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas. For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}. They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites.

For this, we made new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \cite{GomesJunior2015}. We aim to eliminate the systematic errors in the JPL ephemeris pointed out by these authors. %For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.
For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Center and observations from Flagstaff. With this, we predicted stellar occultation for the 9 objects.

Using the derived ephemerides and the UCAC4 catalogue we managed to identify \noccs candidate stellar occultations between January 2016 and December 2020 for the 9 irregular satellites studied here.

Phoebe, being the most studied object with a good measured size, can be used to calibrate and evaluate the technique for similar objects. Up to date, no observation of a stellar occultation by an irregular satellite was published. Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have discouraged earlier tries. But, in fact, given their relatively closer distances as compared to TNOs and Centaurs, and considering the precision of our ephemeris and of star positions, we can now reliably predict the exact location and instant where the shadow of the occultation will cross the Earth.

\begin{table}
\caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
\begin{center}
\begin{tabular}{lccc}
\hline  \hline
\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite  & mas$^ {a}$  & km & Ref. \tabularnewline
\hline
Ananke & 8 & 29 & 1 \tabularnewline
Carme & 13 & 46 & 1 \tabularnewline
Elara & 24 & 86 & 1 \tabularnewline
Himalia & 41 & $(150\times120) \pm 20^{b}$ & 2 \tabularnewline
Leda & 5 & 20 & 1 \tabularnewline
Lysithea & 10 & 36 & 1 \tabularnewline
Pasiphae & 17 & 62 & 1 \tabularnewline
Sinope & 10 & 37 & 1 \tabularnewline
\hdashline
Phoebe & 32 & $212 \pm 1.4^{b}$ & 3 \tabularnewline
%\hdashline
%Nereid & 15 & $340 \pm 50^{c}$ & 4 \tabularnewline
%Triton & 124 & $2707 \pm 2.0^{c}$ & 5 \tabularnewline
\hline
\end{tabular}
\end{center}
References: (1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}.\\
$^{a}${Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
$^{b}${From Cassini observations.}
\par
\end{table}

%Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved. In contrast to TNOs, the irregular satellites have much better ephemeris because the orbits of their host planets are better  known, their observational time span is much wider and covers many orbital periods. Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers. These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites.Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as those by TNOs.

\section*{Orbit computations} \label{Sec: integration}

\cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.

Here we compute new orbits based on the observations published in \cite{GomesJunior2015}. First, because the reduction was made with a consistent and precise stellar catalogue and with a robust astrometry \citep[PRAIA,][]{Assafin2011}. Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets. For these reasons, only this set of positions was used for the satellites of Jupiter.

Due to the context of this work regarding to stellar occultations, the orbit fitting procedures used aimed primarily to derive precise ephemerides for the near future. The procedures allow for the improvement of orbits as new observations are added.

\subsection*{Special tailored Ephemerides (STE) for Jupiter irregular satellites}

The last observations used to develop JPL current ephemeris of the irregular satellites of Jupiter were obtained in 2012 \citep{Jacobson2012}. As a result, the errors in the JPL ephemeris for the current epoch are large enough to prevent accurate predictions of stellar occultations without any corrections.

%Our numerical model describes the dynamical evolution of the irregular satellites of Jupiter in a jovicentric reference frame. The satellites are submitted to the influence of the Sun and the rest of the solar system, as well as that of the Galilean satellites and the first harmonics of Jupiter's gravity field. The axes of the reference frame are those of the ICRS. 
%
%We use the following notations:
%\begin{itemize}
%\item $i$ and $l$ one of the irregular satellites of Jupiter
%\item $J$ Jupiter 
%\item $j$ another body of the Solar System
%\item $M_j$ the mass of the $j$-th body, not an irregular satellite
%\item $m_i$ the mass of the irregular satellite $i$
%\item $\vec{r_i}$ the position of the $i$-th body with respect to the barycentre of Jupiter System
%\item $r_{ij}$ the distance between bodies $i$ and $j$  
%\item $R_J$ the radius of Jupiter
%\item $J_n$ the dynamic polar oblateness of the n-th order for Jupiter's gravity field
%\item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
%\item $\Phi_i$ is the inclination of the $i$-th satellite with respect to Jupiter's equator.
%\end{itemize}
%
%For an irregular satellite $i$, under the gravitational influence of Jupiter, the $\mathcal{N}-1$ other irregular satellites, the regular Jovian satellites and the rest of the Solar System ($\mathcal{N}$ bodies), the equation of motion is:
%\begin{equation}\begin{array}{ll}
%
%\ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N'}Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
%&\displaystyle -\sum_{j=1}^\mathcal{N}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3} \right)\\
% & \displaystyle +GM_J \nabla U_{\bar{l}\hat{J}} -\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}
%\end{array}
%\label{Eq:1}
%\end{equation}
%where the last term in the brackets and the last term in Eq. \ref{Eq:1} represent undirect perturbations. The oblateness potential seen by the body $i$ because of Jupiter is:
%\begin{equation}\begin{array}{ll}
%
%U_{\bar{l}\hat{J}}=&\displaystyle -\frac{R_J^2 J_2}{r_{iJ}^3}\left(\frac{3}{2}\sin^2 \Phi_i-\frac{1}{2}\right)\\ &\\ & 
%\displaystyle-\frac{R_J^4 J_4}{r_{iJ}^5}\left(\frac{35}{8}\sin^4 \Phi_i-\frac{15}{4}\sin^2 \Phi_i+\frac{3}{8}\right)\\
%& \\
%&\displaystyle-\frac{R_J^6 J_6}{r_{iJ}^7}\left(\frac{231}{16}\sin^6 \Phi_i-\frac{315}{16}\sin^4 \Phi_i+\frac{105}{16}\sin^2 \Phi_i-\frac{5}{16}\right)
%
%\end{array}
%\end{equation}
%
%The expressions of $\nabla_lU_{\bar{l}\, \hat{i}}$ and $\nabla_i U_{\bar{i}\, \hat{l}}.$ have been developed in \citet{Lainey2004}. The equations of motion are integrated with the numerical integrator RADAU \citep{Everhart1985}. 
%Our model was fitted to the observations through a least-squares procedure. The satellites were integrated one dynamical family at a time, to gain computing time, while losing minimum precision. Indeed, the interactions between satellites not belonging to the same dynamical family are negligible considering the short timespan of our integration. 
%
%The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.
%
%\begin{table*}
%\scalefont{0.8}
%\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.
% }\label{Tab: sat_ell}
%\begin{center}
%\begin{tabular}{ccccccc}
%\hline\hline
%Satellite & a (km) & e & I$\degr$ & $\Omega\degr$ & $\omega\degr$ & $v\degr$ \\ 
%\hline
%Himalia &   11372100 $\pm$ 500    &    0.166 $\pm$ 0.002      &   45.14 $\pm$ 0.15      &   39.77 $\pm$ 0.19      &   351.48 $\pm$ 0.46      &   97.35 $\pm$ 0.48    \\
%Elara &   11741170 $\pm$ 690  &      0.222 $\pm$ 0.002      &   28.64 $\pm$ 0.18      &   68.42 $\pm$ 0.43      &   179.82 $\pm$ 0.56      &   339.08 $\pm$ 0.82  \\
%Lysithea &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
%Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
%Pasiphae &  23425000  $\pm$ 5000    &     0.379  $\pm$ 0.001       &   152.44 $\pm$ 0.10      &   284.59 $\pm$ 0.21      &   135.96 $\pm$ 0.19      &   236.97 $\pm$ 0.16 \\
%Sinope &   22968800 $\pm$ 5200   &     0.316 $\pm$ 0.002      &   157.76 $\pm$ 0.12      &   256.62 $\pm$ 0.55      &   298.38 $\pm$ 0.55      &   167.57 $\pm$ 0.19    \\
%Carme &   24202924 $\pm$ 4800      &  0.242 $\pm$ 0.001      &   147.13 $\pm$ 0.10      &   154.01 $\pm$ 0.25      &   47.90 $\pm$ 0.29      &   234.41 $\pm$ 0.19  \\
%Ananke &  21683800  $\pm$ 7200  &     0.380 $\pm$ 0.002      &   172.29 $\pm$ 0.20      &   56.9 $\pm$ 1.2      &   123.3 $\pm$ 1.2      &   231.24 $\pm$ 0.21  \\
%%Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\
%\hline
%\end{tabular} 
%\end{center}
%\begin{flushleft}
%\textbf{Notes}: a: semimajor axis; e : excentricity; I: inclination relative to the equatorial reference plane J2000; $\Omega$: longitude of the ascending node; $\omega$: argument of periapsis; $v$: true anomaly.
%\end{flushleft}
%\end{table*}

All the orbits determined for the satellites show satisfying residuals. The residuals are smaller than those obtained with JPL ephemeris, which was expected because the accuracy of an ephemeris decreases when we get further from the time of observations. The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations. If that were the case, our ephemeris would diverge too quickly to be of any use. JPL ephemerides are fitted over all the available observations. As a result, they will diverge less quickly than our own. Though they are no longer precise enough for our use, they remain a precious reference to identify whether our own model presents a quick divergence.

We compared our ephemeris to the JPL for all the Jupiter satellites we fitted, until 2018. For instance, the divergence between 2015 and 2018 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.

Fig. \ref{Fig: JPL-STE} displays the offsets of the positions published by \cite{GomesJunior2015} for the satellite Carme in declination relative to our ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \cite{GomesJunior2015} are reduced with our ephemeris, as expected.

\begin{figure}
\includegraphics[width=16.0cm]{figures/Carme_ephemeris.eps}
\caption{Offsets in declination of the positions published by \protect\cite{GomesJunior2015} for Carme. The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the JUP300 JPL ephemeris and the green dot relative to \protect\cite{Emelyanov2008}. As expected, the ephemeris systematic errors pointed out by \protect\cite{GomesJunior2015} are reduced with the STE ephemeris. \label{Fig: JPL-STE}}
\end{figure}

The obtained ephemeris is hereafter refered as STE, for special-tailored ephemeris.

\subsection*{Phoebe's ephemeris}

For the specific case of Phoebe, the ninth satellite of Saturn, we have updated the ephemeris published in \cite{Desmars2013}. The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter. The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center\footnote{\url{http://lnfm1.sai.msu.ru/neb/nss/bsapoouf.htm}}), and observations from Flagstaff \citep{NOFS} between 2012 and 2014. It represents a total number of 5886 observations from 1898 to 2014. In contrast, in \cite{Desmars2013} was used 3367 observations from 1898 to 2012. It represents an increase of almost 75\% in the number of observations.

In Fig. \ref{Fig:eph-Phoebe} we compare our ephemeris (PH15) with the SAT375 JPL\footnote{Jacobson, R.A. 2015-Feb-27. "Satellite Ephemeris: SAT375", JPL Satellite Ephemeris File Release, \url{ftp://ssd.jpl.nasa.gov/pub/eph/satellites/nio/LINUX_PC/sat375l.txt}} ephemeris. The difference between them is smaller than 30 mas (< 10 mas in Declination). This difference is smaller than the apparent diameter of Phoebe (see Table \ref{Tab: satellite-diameter})

\begin{figure}
\begin{centering}
\includegraphics[width=16.0cm]{figures/Phoebe.eps} 
\caption{Comparison between the PH15, SAT375 JPL and \protect\cite{Emelyanov2007} ephemeris for the satellite Phoebe.}
\label{Fig:eph-Phoebe}
\end{centering}
\end{figure}

%\subsection{Triton's and Nereid's ephemeris}
%
%Triton is an uncommon satellite. Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones. Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but still studied together by many authors \citep{Sheppard2005, Jewitt2007}. Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}. However, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
%
%For this satellite, we used the most recent ephemeris published by \cite{Emelyanov2015} which is based on more observations than previous ephemeris and covers 4 years more than the ephemeris of \cite{Jacobson2009}. In Fig. \ref{Fig:eph-Triton} we compare it with the ephemeris published in \cite{Zhang2014} and the NEP081 JPL ephemeris \citep{Jacobson2009}. The offsets between the three ephemeris are smaller than 20 mas for the period 2015-2021. This value is much smaller than the apparent size of Triton (124 mas, see Table \ref{Tab: satellite-diameter}), which indicates a good probability of observing an event by this object.
%
%\begin{figure*}
%\begin{centering}
%\subfigure[NEP081 JPL and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
%\subfigure[\protect\cite{Zhang2014} and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/Zhang-EME_Triton.eps} \label{Fig:eph-Triton-zhang}}
%\caption{Comparison between the \protect\cite{Emelyanov2015}, \protect\cite{Zhang2014} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Triton.}
%\label{Fig:eph-Triton}
%\end{centering}
%\end{figure*}
%
%For Nereid, we used the ephemeris published by \cite{Emelyanov2011}, which uses more recent observations than the JPL ephemeris published by \cite{Jacobson2009}. The comparison between the two ephemeris can be seen in Fig. \ref{Fig:eph-Nereid}. The right ascension offsets between them are smaller than 60 mas and the declination ones are smaller than 15 mas. For Nereid, ephemeris errors in right ascension translate to uncertainties in the central instant of a stellar occultation, so offsets of 60 mas are not critical. The declination ephemeris offsets are of the order of the apparent diameter (15 mas, see Table \ref{Tab: satellite-diameter}), resulting in fair chances (50\%) that the shadow crosses the expected Earth latitude predicted for an occultation. 
%
%\begin{figure}
%\begin{centering}
%\includegraphics[width=8.8cm]{figures/JPL-EME_Nereid.eps}
%\caption{Comparison between the \protect\cite{Emelyanov2011} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Nereid.}
%\label{Fig:eph-Nereid}
%\end{centering}
%\end{figure}
%
%\cite{GomesJunior2015} published 902 new positions for Nereid from observations between 1992 and 2014. The development of a better ephemeris for Nereid using these new positions in the context discussed here of near future stellar occultations is desirable, but is out of the scope of this paper.

\section*{Prediction of stellar occultations} \label{Sec: predictions}

\subsection*{Candidate events}

The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the ephemeris presented in the previous section. The search for stellar candidates follows the same procedure as presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.

We predicted occultations for the 8 major irregular satellites of Jupiter,  Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope, and for Phoebe of Saturn.

A total of \noccs events were identified between January 2016 and December 2020. In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite. It is possible to see an increase in the number of events found for Phoebe in 2018 and for the satellites of Jupiter in 2019-2020. This is because at that periods these satellites will cross the apparent galactic plane. We call attention that about 10\% of the events will involve stars brighter than magR=14 (and almost 25\% brighter than magR=15), which helps the attempt of amateur observers.

\begin{table}[h]
\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2020.}
\begin{centering}
\begin{tabular}{lcccccc}
\hline  \hline
%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite  & 2016 & 2017 & 2018 & 2019 & 2020 & Total \tabularnewline
\hline
Ananke & 12 & 16 & 49 & 359 & 187 & 623 \tabularnewline
Carme & 20 & 14 & 30 & 369 & 220 & 653 \tabularnewline
Elara & 14 & 16 & 33 & 305 & 193 & 561 \tabularnewline
Himalia & 15 & 12 & 54 & 257 & 230 & 568 \tabularnewline
Leda & 8 & 24 & 38 & 362 & 208 & 640 \tabularnewline
Lysithea & 16 & 11 & 35 & 330 & 212 & 604 \tabularnewline
Pasiphae & 20 & 19 & 44 & 362 & 206 & 651 \tabularnewline
Sinope & 15 & 21 & 34 & 356 & 256 & 682 \tabularnewline
\hdashline
Phoebe & 32 & 98 & 238 & 79 & 13 & 460 \tabularnewline
%\hdashline
%Nereid & 11$^{a}$ & 1 & 1 & 0 & 0 & 13 \tabularnewline
%Triton & -- & 1 & 1 & 0 & 0 & 2 \tabularnewline
\hline
\end{tabular}
\par \end{centering}
\end{table}

%Table \ref{Tab: sample-cds} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps. Since these objects are very small, the duration of each event is a few seconds. All the occultation tables and maps will be publicly available at the CDS.
In Fig. \ref{Fig: ocultacao} we show an example of an occultation map. This is an occultation by Elara that will happen in February 21, 2017. This event can be observed from Australia and it is one of the best opportunities for this object due to the slow velocity of the event and it involves a bright star (MagR*=12.4).

%\begin{table*}
%\caption{\label{Tab: sample-cds} A sample of stellar occultation predictions for Pasiphae}
%\begin{center}
%\begin{tabular}{cccrcccrcrr}
%\hline
%\hline
%d m Year~~~~h~~m~~~s & RA~~~(ICRS)~~~Dec & C/A & P/A & $\nu$ & {\it D} & $R^*$ & $\lambda$ & LST & $\mu_{\alpha*}$ & $\mu_{\delta}$ \tabularnewline
%\hline
%% & \multicolumn{2}{c}{Mean errors} & Nr & Nr &  UCAC4       \tabularnewline
%09 04 2016 03:58:19. & 11 14 36.7707 +07 39 20.7610 & 1.003 &  17.9 & -12.88 &  4.54 & 14.9 & 271. & 22:03 &  12. & -33. \tabularnewline 
%13 06 2016 00:16:12. & 11 12 48.5020 +07 06 43.3520 & 0.661 &  30.0 & +14.32 &  5.50 & 13.9 & 262. & 17:45 &  -1. &   1. \tabularnewline 
%27 06 2016 13:56:09. & 11 18 03.4160 +06 23 45.1940 & 1.707 &  28.0 & +20.29 &  5.74 & 11.7 &  44. & 16:53 &   4. & -10. \tabularnewline 
%18 07 2016 15:07:24. & 11 28 15.5076 +05 05 31.8060 & 0.942 &  26.7 & +27.80 &  6.05 & 14.0 &   8. & 15:40 &   4. &   4. \tabularnewline 
%22 07 2016 16:15:07. & 11 30 30.4310 +04 48 43.4340 & 0.644 & 206.5 & +29.04 &  6.11 & 14.6 & 348. & 15:27 &  23. & -24. \tabularnewline 
%24 07 2016 01:37:34. & 11 31 17.8471 +04 42 49.0540 & 0.029 & 206.6 & +29.46 &  6.12 & 15.1 & 206. & 15:22 &   2. &  -8. \tabularnewline 
%24 07 2016 17:37:18. & 11 31 40.7472 +04 39 57.5060 & 0.840 &  26.5 & +29.66 &  6.13 & 14.9 & 326. & 15:20 & -11. &  -1. \tabularnewline 
%\hline
%\end{tabular}
%\end{center}
%\begin{flushleft}
%\textbf{Notes}. Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - at the central instant of occultation (corrected by proper motions); C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees (zero at north of the star, increasing clockwise); $\nu$: velocity in the plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
%\end{flushleft}
%\end{table*}

\begin{figure}
\begin{centering}
\includegraphics[width = 16.0cm]{figures/Elara_2017-02-21T15:28:40.png}   
\caption{Occultation map for Elara. % regarding to the first event sampled in Table \ref{Tab: occ-list}.
The central red dot show the geocentric closest approach of the shadow. The small ones shows the center of the shadow separated by 60s. The lines show the path of the shadow over the Earth. The shadow moves from right to left.
\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for a central observation; C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; vel: velocity of event in km/s; Delta: Geocentric distance to the occulting object in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$; long: east longitude of subplanet point in degrees, positive towards east.}
\label{Fig: ocultacao}
\end{centering}
\end{figure}

%The first preliminary catalogue version of the ESA astrometry satellite GAIA \citep{deBruijne2012} is expected to be released up to the end of 2016 (The catalogue with five-parameter astrometric solutions is up to the end of 2017). The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris. Astrometric reduction of observations published in \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions will be improved. Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish predictions up to the end of 2020. In the GAIA era, the occultations predicted will be updated.

\subsection*{Robustness of predictions} \label{Sec: testes}

%Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
In contrast to TNOs, the irregular satellites have much better ephemeris because the orbits of their host planets are better  known, their observational time span is much wider and covers many orbital periods. Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers. These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites. Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as those by TNOs.

%Observing a stellar occultation demands a great effort. And, in our case, the shadow covers a very restricted area on Earth because of the size of the irregular satellites. Since no stellar occultation by an irregular satellite was observed up to date, and since we want to be sure that we can start observational campaigns with reasonable chances of success,

To check the availability of observing an occultations we tested the robustness of a prediction for a large target. The test design consisted in observing the object and star to be occulted near the date of the event predicted when the two objects were present in the same field of view (FOV), close to each other. Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions %\citep[and references therein]{Peng2008}. 
The relative positions of the star and satellite were used to check the original prediction. Notice that in the test we did not attempt to observe any actual occultation. The test could be performed at any site, regardless of the Earth location where the occultation would in fact be visible.

We tested the occultation by Himalia predicted to occur on March 3, 2015. The shadow would cross the northern part of South America. For the event, four situations were considered:
\begin{enumerate}
\item Our nominal, published prediction with the STE ephemeris, and the nominal UCAC4 position of the star;
\item Prediction with the JPL ephemeris and the nominal UCAC4 position of the star;
\item From star and satellite offsets calculated from observations made a few days before the occultation when the objects were very separated (different FOVs);
\item Same as 3 but with the star and the satellite close in the same FOV.
\end{enumerate}

>>>>

\documentclass[12pt,a4paper]{report}
\usepackage[brazilian, english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amsthm,amsfonts,amssymb,textcomp}
%\usepackage{latexsym}
\usepackage{graphicx}
\graphicspath{{figuras}}
\usepackage{subfigure}
\usepackage{float}
\usepackage{longtable}
\usepackage{color}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage[breaklinks=true]{hyperref}
\usepackage[comma,authoryear]{natbib}
\usepackage[nonumberlist]{glossaries}
\usepackage{arydshln}
\usepackage{footnote}
\usepackage{longtable}
\usepackage[small,bf,singlelinecheck=off]{caption}
\usepackage[left=3cm,right=2cm,top=3cm,bottom=2cm]{geometry}
%\usepackage[alf]{abntcite}

%%% newcommand %%%%%%%%%%%%%%%%%
\newcommand{\PE}{Perkin-Elmer }
\newcommand{\BC}{Boller \& Chivens }

@@

%\input{glossario}
\author{Altair Ramos Gomes Júnior}
\title{Astrometry of the Neptune-Triton System}
\begin{document}

@@

\section*{Chromatic Refraction}

Since the colors of Neptune and Triton are different, it is expected that their positions have influence of chromatic refraction with different intensities. The apparent position of Neptune, which is more blue than Triton, would be more shifted towards the zenith than the Triton's position.

To test the effects of chromatic refraction I looked for nights with many observations distributed over the night. I selected two of them (August 22, 2004 and September 25, 2011) with more than 4 hours between the first and last observations. In the night of 2004 it was used a filter called "Dark" which we suppose it means no filter, however we are no sure so I classified it as unknown. In the night of 2011 it was used a I filter.

In Fig \ref{Fig:refraction} it is plotted the difference in the positions of Triton and Neptune compared to the difference of their ephemeris in RA and DEC for both nights.

\begin{figure}[h]
\includegraphics[width=16.0cm]{plot_hour_gr1.png} 
\caption{Distribution of the difference of offsets in the sense Triton - Neptune for the nights August 22, 2004 and September 25, 2011.}
\label{Fig:refraction}
\end{figure}

Teoretically, the difference in the positions of the objects in the sense Triton - Neptune compared to the difference in the ephemeris over a night would cause the following effects.

\begin{itemize}
\item RA: the difference in the offsets would be positives in the East side of the sky, negative in the West side and zero in the culmination, with the assumption that only chromatic refraction affects the offsets.
\item DEC: the difference in the offsets in the culmination would be positive (Neptune is in the North side in both nights for the site). Farther from the culmination the difference in the offsets would be more positive.
\end{itemize}

Fig \ref{Fig:refraction} clearly shows that that the chromatic refraction is affecting the offsets confirming our expectation.

\begin{table*}[h]
\centering
\caption{Difference in the offsets of Triton-Neptune of the nights used as tests of chromatic refraction with sigma-clipping procedure}
\label{Tab:dados}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline 
Night & N & $\mu$ RA & $\mu$ DEC & $\sigma$ RA & $\sigma$ DEC & Ncat\\
\hline
2004-08-22 & 29 & 2 & 45 & 34 & 28 & 13 \\ 
2011-09-25 & 106 & -20 & 67 & 11 & 11 & 17\\
\hline 
\end{tabular}
\\N: Number of positions after sigma-clipping procedure. Ncat is the number UCAC4 stars used in the reductions.
\end{table*}
